Synthesis of Linear Functional Programs

### 2021-05-07 article/article.tex
tweaks

\newcommand{\lolli}{\multimap}
\newcommand{\tensor}{\otimes}
\newcommand{\one}{\mathbf{1}}
\newcommand{\bang}{{!}}

@@

% General motivation?
Program synthesis is an automated or semi-automated process of
deriving a program (i.e.~generating code) from a high-level specification.
Specifications can take many forms (e.g.  polymorphic refinement
types~\cite{DBLP:conf/pldi/PolikarpovaKS16},
examples~\cite{DBLP:conf/popl/FrankleOWZ16}, graded linear
types~\cite{DBLP:conf/lopstr/HughesO20}, ...).  In this work we
explore type-based synthesis where specifications take the form of
linear types.\todo{O que é um linear type?}
%
The chosen specification form dictates the synthesis process, and
different ones might be approached in completely different
ways.\todo{Esta frase não parece ter grande utilidade. O que queres
  dizer aqui?}
%
Regardless of the kind of specification, program synthesizers have to
deal with two main inherent sources of complexity: searching over a
vast space of (potentially) valid programs, and interpreting user
intent.
%
Synthesis has many motivations, it can be seen as a means to
improve a programmer's productivity and/or program correctness
(i.e. through suggestion and/or autocompletion), or as a tool to
automate certain parts of the programming process (e.g. in the same
way AI might generate some boilerplate text to auxiliate a journalist,
rather than to replace), just to name a few examples.
\todo[inline]{Acho que estes 3 ``paragrafos'' deviam estar por ordens diferentes.}

% Type based synthesis
Type-based synthesis takes a type and produces a program with that
given type.  Types can be further explored to give the programmer more
expressiveness (e.g. by adding constraints??), and are called rich
types.  -- Gostava de ver com o prof aqui. [Estive a procurar e fiquei
um pouco confuso. polimorfismo é rich types?  dependent types are rich
types? aqueles constraints são os dependent types pelo que
percebi... mas n me parece correto. arithmetic types? temos de
rever. Por agora vou por apenas os tópicos que o professor escreveu]
Languages with richer type systems allow for more precise types. This
precision can statically eliminate various kinds of logical errors by
making certain invalid program states ill-typed (e.g. a ``null aware''
type system, will ensure you cannot dereference a null-pointer).
However, they can also be a burden -- the whole point of these type
systems is to ensure that ``less'' things are well-typed, and
sometimes it's hard to convince the checker.  Regarding the
challenges, type-based synthesis leverages rich types as a way of
pruning the search space, and by using types gives the user as a more
``familiar'' specification.

This work explores the process of synthetising linear functional
programs from types based in linear logic (linear types), leveraging
the Curry-Howard correspondence.  Said correspondence states that
propositions in a logic have a direct mapping to types, and well typed
programs also have a one to one correspendence to proofs of those
propositions.  As such, having a type be a proposition in linear
logic, we can relate a proof of that proposition directly to a linear
functional program — finding a proof is finding a program with that
type, so we formulate the synthesis as proof search in linear logic,
which makes to us available a lot of related technology from the proof
search literature.  Linear types differ from usual types for they
constrain resource usage in programs by \emph{statically?} limiting
the number of times certain resources can be used during their
lifetime.  They can be applied to resource-aware programming such as
concurrent programming (e.g. session types??), and to
memory-management (e.g. Rust) [Quero reescrever a frase acima, não
gosto de como falamos de session types e rust em dois exemplos
paralelos visto que não são coisas bem comparáveis. Revemos na
reunião.]

 % Goals
In the end, we intend to be able to do full and partial synthesis of
well-typed programs -- Full synthesis being the production of a
function (or set of) satisfying the specification; Partial synthesis
being the ``completion'' of a partial program (i.e. a function with a
\emph{hole} in it) -- Starting from a small core linear --
i.e. \emph{resource-aware} functional language, and building up to
recursive types/functions.  Eventually, possibly, enhancing the
synthesis process with type refinements, and making it interactive --
i.e. the synthetiser offers the user ``choices'' mid-process.  [Devia
estar a escrever isto eu ahah?]The evaluation will be done through
expressiveness benchmarks (i.e. can we synthesize ``X''?); time
measurements (i.e. how fast can we synthesize ``X''?'); and so on.

@@

% Linear Logic
Linear logic \cite{DBLP:journals/tcs/Girard87} can be thought of as a logic of resources.
Where we were previously able to utilize a variable i.e. a hypothesis as many times as we wanted,
now we'll keep track of the amount available (the ``resources''), and guarantee that every resource is used exactly once, hence the name ``Linear Logic''.
It comes that connectives must be decomposed to get more primitive ones that reflect our resource aware system.
So what would be the $\wedge$ (and) in classical logic, decomposes into its multiplicative

>>>>

\newcommand{\lolli}{\multimap}
\newcommand{\tensor}{\otimes}
\newcommand{\one}{\mathbf{1}}
\newcommand{\bang}{{!}}
\newcommand{\mypara}[1]{\paragraph{\textbf{#1}.}}

@@

% General motivation?
Program synthesis is an automated or semi-automated process of
deriving a program (i.e.~generating code) from a high-level
specification.  
%
Synthesis can be seen as a means to improve a programmer's
productivity and/or program correctness (i.e. through suggestion
and/or autocompletion), or as a tool to automate certain parts of the
programming process (e.g. in the same way AI might generate some
boilerplate text to aid a journalist, rather than to replace the
journalist outright), just to name a few examples.

Specifications can take many forms (e.g.  polymorphic refinement
types~\cite{DBLP:conf/pldi/PolikarpovaKS16},
examples~\cite{DBLP:conf/popl/FrankleOWZ16}, graded linear
types~\cite{DBLP:conf/lopstr/HughesO20}, ...).  Regardless of the kind
of specification, program synthesizers have to deal with two main
inherent sources of complexity: searching over a vast space of
(potentially) valid programs, and interpreting user intent.
%
In this work we will explore type-based synthesis where specifications
take the form of linear types, which are types that limit resource
usage in programs.  Concretely, we will explore the proof-theoretic
foundations of linear types via the Curry-Howard correspondence with
linear logic, viewing program synthesis under the lens of (linear
logic) bottom-up proof search.
%

%

\mypara{Type-Based Synthesis}
% Type based synthesis
Type-based synthesis uses types as a form of program specification,
automatically or semi-automatically producing a program expression
with the given type and so matching the specification.
%
In order to make type-based specifications more expressive, type-based
synthesis frameworks use rich type systems such as refinement
types~\cite{} and polymorphic types~\cite{}.
\todo[inline]{[Estive a procurar e fiquei
um pouco confuso. polimorfismo é rich types?  dependent types are rich
types? aqueles constraints são os dependent types pelo que
percebi... mas n me parece correto. arithmetic types? temos de
rever. Por agora vou por apenas os tópicos que o professor escreveu]
-- Não há propriamente uma definição rigorosa de ``rich types'',
geralmente tudo o que vá além do tipico pode ser qualificado como
rico.}

Richer type systems allow for more precise types, which can
statically eliminate various kinds of logical errors by making certain
invalid program states ill-typed (e.g.,~a ``null aware'' type system,
will ensure at compile-time that you cannot dereference a
null-pointer). However, they can also be a burden -- the whole point
of these type systems is to ensure that \emph{fewer} programs are
deemed well-typed, which can pose additional challenges to the
development process. Type-based synthesis leverages rich types as a
way of pruning the search space, and by using types gives the user as
a more ``familiar'' specification. For instance, the type
$\mathsf{Int} \rightarrow \mathsf{Int} \rightarrow \mathsf{Int}$
specifies a (curried) function that takes two integers and produces an
integer. Viewed as a specification, it is extremely imprecise (there
are an infinite number of functions that satisfy this specification).
However, the richer type $(x{:}\mathsf{Int}) \rightarrow
(y{:}\mathsf{Int}) \rightarrow \{z{:}\mathsf{Int} \mid z = x+y\}$
very precisely specifies a function that takes integer arguments $x$
and $y$ and returns an integer $z$ that is the sum of $x$ and $y$.

%
This work explores the process of synthesizing linear functional
programs from types based on linear logic propositions (i.e.~linear
types) by leveraging the Curry-Howard correspondence.  The
correspondence states that propositions in a logic have a direct
mapping to types, and well-typed programs correspond to to proofs of
those propositions.  As such, having a type be a proposition in linear
logic, we can relate a proof of that proposition directly to a linear
functional program — finding a proof is finding a program with that
type. Thus, we formulate synthesis as proof search in linear logic,
which allows us to inform our work with approaches from the proof
search literature.  Linear types differ from more traditional types in
that they constrain resource usage in programs by \emph{statically}
limiting the number of times certain resources can be used during
their lifetime.  They can be applied to resource-aware programming
such as concurrent programming (e.g. session types for message passing
concurrency~\cite{}), and to memory-management (e.g.~Rust's ownership
types).

\mypara{Goals.}
 % Goals
In the end, we intend to be able to do full and partial synthesis of
well-typed programs. Full synthesis consists of the production of a
function (or set of) satisfying the specification; Partial synthesis
is the ``completion'' of a partial program (i.e. a function with a
\emph{hole} in it). The work will start from a small core linear --
i.e. \emph{resource-aware} functional language, building up to
recursive types/functions, with potential other avenues of further
extension. 
%
We will evaluate the work through
expressiveness benchmarks (i.e. can we synthesize ``X''?) and time
measurements (i.e. how fast can we synthesize ``X''?').

@@

% Linear Logic
\subsection{Linear Logic}

Linear logic \cite{DBLP:journals/tcs/Girard87} can be seen as a
resource-aware logic, where propositions are interpreted as resources
that are consumed during the inference process.  Where in standard
propositional logic we are able to use an assumption as many times as
we want, in linear logic every resource (i.e., every assumption) must
be used \emph{exactly once}, or \emph{linearly}. This usage
restriction gives rise to new logical connectives, based on the way
the ambient resources are used. For instance, conjunction, usually
written as $A\wedge B$, appears in two forms in linear logic:
multiplicative or simultaneous conjunction (written $A\tensor B$); and
additive or alternative conjunction (written $A\with
B$). Multiplicative conjunction denotes the simultaneous availability
of resources $A$ and $B$, requiring both of them to be
used. Alternative conjunction denotes the availability of $A$ and $B$,
but where only one of the two resources may be used. Similarly,
implication becomes into linear implication, written $A\lolli B$,
denoting a resource that will consume (exactly one) resource $A$ to
produce a resource $B$.
